---
title: "NBA data"
author: "Walter Mymala"
date: "7/30/2021"
output: html_document
---

```{r}

devtools::install_github("abresler/nbastatR")
library(tidyverse)
library(dplyr)
library(ggplot2)
library(nbastatR) ## data of every NBA player EVER
lsf.str("package:nbastatR")
```

```{r}
# avoid: "size of the connection buffer (131072) was not large enough"
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 2) 
assign_nba_players()
```

players_careers()
• dataPlayerCareerTotalsAllStarSeason
• dataPlayerCareerTotalsPostSeason
• dataPlayerCareerTotalsRegularSeason
• dataPlayerSeasonRankingsPostSeason
• dataPlayerSeasonRankingsRegularSeason
• dataPlayerSeasonTotalsAllStarSeason
• dataPlayerSeasonTotalsPostSeason
• dataPlayerSeasonTotalsRegularSeason

The mode input requires one of the following designations:
Totals
PerGame
Per36

```{r}
# function to get 8 tibbles from above
# players_careers(players = c("LeBron James")) similar thing as below
# kobe_totals <- players_careers(players = c("Kobe Bryant"),
#                modes = c("Totals"))

kobe_totals <- players_careers(players = c("Kobe Bryant"), modes = c("Totals", "PerGame", "Per36"))


## data.frame of active NBA players as of 2020-2021 NBA season
activePlayers <- df_dict_nba_players %>% filter(isActive == "TRUE") %>% select(namePlayer)
activePlayers
```

